Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-07 conclusions.tex
lloid_paper: Polishing pass post-Chad
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@958 ccdd05d5-56c1-461a-bf98-85f6876523b4

The algorithm we described has no intrinsic latency.  However, there are
fundamental and practical latencies associated with the analysis and detection
procedure. For example, the \LIGO{} detectors, data acquisition is synchronized
to a 1/(16~Hz) cadence introducing an up-front latency~\cite{CITE_CDS}.  Data aggregation from
the observatories will travel over various networks, each capable of high
bandwith but perhaps only modest latency.  This could amount to a similar
latency of $\sim$100 ms.  Lastly, unless a real-time infrastructure is adopted
post data acquisition, it is likely that there will be an inherent latency
introduced by such infrastructure.  We have shown a prototype implementation
using \gstlal\ that is capable of order seconds latency. In our opinion,
significant work would have to be done in order to improve upon this number.
For example, a tighter integration of analysis and data acquisition would be
beneficial. This should be considered for third-generation detector design.

Although we have demonstrated a feasible method for advance detection we have
not explored the accuracy of sky localization that is possible before merger.
Ref.~\cite{Fairhurst2009} discusses some of the theoretical prospects for early sky
localization.  Our future work will explore the prospects of early warning
detection with realistic simulations of binary mergers using the infrastructure
and techniques described here.

>>>>

The algorithm we described has no intrinsic latency.  However, there are
fundamental and practical latencies associated with the analysis and detection
procedure. For example, the \LIGO{} detectors, data acquisition is synchronized
to a 1/(16~Hz) cadence, introducing an up-front latency~\cite{CITE_CDS}. Data
aggregation from
the observatories will travel over various networks, each capable of high
bandwith but perhaps only modest latency.  This could amount to a similar
latency of $\sim$100 ms.  Lastly, unless a real-time infrastructure is adopted
post data acquisition, it is likely that there will be an inherent latency
introduced by such infrastructure.  We have shown a prototype implementation
of \lloid{} using \gstlal\ that is capable of order seconds latency. Based
on our experience, significant work would have to be done in order to improve
upon this number, for example, more tightly integrating analysis and data
acquisition. This should be considered for third-generation detector design.

Although we have demonstrated a feasible method for advance detection we have
not explored the accuracy of sky localization that is possible before merger.
Ref.~\cite{Fairhurst2009} discusses some of the theoretical prospects for early sky
localization.  Our future work will explore the prospects of early-warning
detection with realistic simulations of binary mergers using the infrastructure
and techniques described here.

### 2011-06-07 implementation.tex
lloid_paper: Polishing pass post-Chad
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@958 ccdd05d5-56c1-461a-bf98-85f6876523b4

The choice of templates and the \SVD\ can be done in advance.  New templates
and the \SVD\ can be computed asynchronously using updated spectrum estimates
as they are available.

@@

The \lloid\ algorithm could be used in a true sample-in-sample-out real-time
system.  However, such a system would likely require integration directly into
the data acquisition and storage system of the gravitational-wave
observatories.  A slightly more modest goal is to leverage existing low
latency, but not real-time, signal processing infrastructure in order to
implement the \lloid\ algorithm.  For the near-term this should be a viable 
solution for ultra-low latency searches with order seconds of intrinsic latency.

@@

The early-warning \SNR\ for each time slice is formed by accumulating interpolated early
warning \SNR\ from the subsequent time slice.  This process continues until we have 
worked our way to the \SNR\ of the original templates at the full sample rate $f^0$.  In 
this way, the \lloid\ algorithm and this implementation leads to a simple early warning
pipeline.

>>>>

@@

The \lloid\ algorithm could be used in a true sample-in-sample-out real-time
system.  However, such a system would likely require integration directly into
the data acquisition and storage system of the gravitational-wave
observatories.  A slightly more modest goal is to leverage existing low-%
latency, but not real-time, signal processing infrastructure in order to
implement the \lloid\ algorithm.  For the near-term this should be a viable 
solution for searches with order seconds of intrinsic latency.

@@

The early-warning \SNR\ for each time slice is formed by accumulating
interpolated early-warning \SNR\ from the subsequent time slice.  This process
continues until we have worked our way to the \SNR\ of the original templates
at the full sample rate $f^0$.  In this way, the \lloid\ algorithm and this
implementation naturally leads to a simple early-warning pipeline.

### 2011-06-07 introduction.tex
lloid_paper: Polishing pass post-Chad
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@958 ccdd05d5-56c1-461a-bf98-85f6876523b4

Fortunately, the morphology of inspiral signals can be exploited to offset some
of the computational complexity of low-latency algorithms.  First, the signals
evolve slowly in frequency, so that they can be broken into contiguous
band-limited time intervals and processed at possibly lower sample rates.
Second, inspiral filter banks consist of highly similar templates, admitting
methods such as singular value decomposition to reduce the number of templates\cite{Cannon:2010p10398}
We will use both aspects to demonstrate that a very
low latency analysis with advance detection of compact binary sources is
possible with current computing resources.  Assuming other technical sources of
latency can be reduced significantly, this should allow the possibility for
prompt alerts to be sent to the astronomical community.

>>>>

Fortunately, the morphology of inspiral signals can be exploited to offset some
of the computational complexity of low-latency algorithms.  First, the signals
evolve slowly in frequency, so that they can be broken into contiguous
band-limited time intervals and processed at possibly lower sample rates.
Second, inspiral filter banks consist of highly similar templates, admitting
methods such as singular value decomposition (\SVD{}) to reduce the number of templates\cite{Cannon:2010p10398}
We will use both aspects to demonstrate that a very
low latency analysis with advance detection of compact binary sources is
possible with current computing resources.  Assuming other technical sources of
latency can be reduced significantly, this should allow the possibility for
prompt alerts to be sent to the astronomical community.

### 2011-06-07 method.tex
lloid_paper: Polishing pass post-Chad
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@958 ccdd05d5-56c1-461a-bf98-85f6876523b4

In this section we describe a decomposition of the \CBC{} waveform space that
reduces time-domain (\TD) filtering cost sufficiently to allow for the
possibility of \earlywarning\ detection with modest computing requirements.  We
expand on the ideas of~\cite{Marion2004, Buskulic2010} that describe a
multiband decomposition of the compact binary parameter space that resulted in
a search with minutes latency in \LIGO{}'s S6 and Virgo's VSR3 joint science
runs~\cite{HugheyGWPAW2011}.  We combine this with the \SVD\ rank-reduction
method described in~\cite{Cannon:2010p10398} that exploits the redundancy of
the template banks.

@@

Here we explore a method for reducing the computational cost of a \TD\ search
for compact binary coalescence.  We will give a truly zero-latency algorithm
that competes in terms of floating point operations per second with the
conventional \FD\ method, which requires a signficant latency in order to be
computationally cheap. Our method, \lloid\footnote{\lloid, The term strictly
means \emph{Low Latency Online Inspiral Detection} and was a code name for
this procedure as it was being developed. It is now known in some GW circles.
We feel that although the name doesn't carry much specific information about
the details of the algorithm it is far too late to do anything about it now.}
involves two transformations of the template waveforms that produce a set of
orthogonal filters with far fewer coefficients than the original templates.

The first transformation is to chop the \TD{} templates up into \emph{time
slices}.  Since each template slice is disjoint in time, the resulting set for
a single template is orthogonal.  Given the chirp-like structure of the
templates, the early time slices have significantly lower bandwidth and can be
safely downsampled.  The downsampling removes a factor of $\sim$100 samples
from the early part of the waveform and allows the filters to be evaluated at
about $\sim$100 times lower rate.  This amounts to more than a factor of 10000
reduction in the floating-point operations per second required to filter the
early part of the waveform.  However, the resulting filters are still not
orthogonal across the mass parameter space, and are in fact highly redundant.
We use the singular value decomposition to produce an orthogonal filter set
from the time-sliced templates~\cite{Cannon:2010p10398}.  We find that this
reduces the number of sample points of the filters by another factor of
$\sim$100.  The combined methods reduce the number of floating point operations
to the level where they are competitive with the conventional high-latency \FD\
matched filter approach.  In the remainder of this section we describe the
\lloid\ algorithm in detail and provide some basic computational cost scaling.

@@

The first step of our proposed method is to divide the templates into
time slices in a \TD{} analogue to the \FD{}
decomposition described in~\cite{Marion2004, Buskulic2010}.
A matched filter is constructed for each time
slice.  The outputs form an ensemble of partial \SNR{} streams.  By linearity,
these partial \SNR\ streams can be suitably time-delayed and summed to
reproduce the \SNR\ of the full template.  To wit,
%
\begin{equation}
\label{eq:time-slices}
h_{i}[k] = \sum_{s=0}^{S-1}
	\cases{%
		h_i^s[k] & for $t^s \leqslant k / f^0 < t^{s+1}$ \\
		0 & otherwise
	}
\end{equation}
\editorial{Drew: Should this be $h_{i}[k]$ or $\rho_{i}[k]$?}
%
for $S$ integers $\{f^0 t^s\}$ such that $0  = f^0 t^0 < f^0 t^1 < \cdots < f^0 t^S = N$.
We will show
in the next section that this, combined with the singular value docomposition,
is sufficient to enable a computationally efficient \TD{} search and
furthermore is an essential part of an \earlywarning\ detection scheme.

@@

The monotonic time-frequency relationship of equation~\eqref{eq:fgw} allows us
to choose time-slice boundaries that require substantially less bandwidth at
early times in the inspiral.  Our goal is to reduce the filtering cost of a
large fraction of the waveform by computing part of the convolution at a lower
sample rate.  Specifically we consider here time slice boundaries with the
smallest power-of-two sample rate that sub-critically samples the time sliced
template.  The time slices consist of the $S$ intervals
$\left[t^0, t^1\right),\, \left[t^1, t^2\right),\, \dots,\, \left[t^{S-1}, t^S\right)$,
sampled at frequencies $f^0,\, f^1,\, \dots,\, f^\mathrm{S-1}$ where $f^0
\geqslant 2 f_\mathrm{hi}$ and $f^{S-1} \geqslant 2 f_\mathrm{low}$.  The time
sliced templates may be downsampled without aliasing, so we define them as
%
\begin{equation}
\label{eq:time-sliced-templates}
h_{i}^{s}[k] \equiv
	\cases{
		h_{i}\!\left[k\frac{f}{f^s}\right] & if $t^s \leqslant k/f^s < t^{s+1}$ \\
		0 & otherwise.
	}
\end{equation}
%
\begin{comment}
An example time-slice design satisfying these constraints for a $1.4 - 1.4 \,
M_{\odot}$ binary is shown in table~\ref{table:time_slices}.
%
\begin{table}[h!]
\begin{minipage}[c]{0.52\textwidth}
\centering
\vspace{0.8cm}
\includegraphics{time_slices.pdf}
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\centering
\input{time_slices.tex}
\end{minipage}
\caption{\label{table:time_slices} Example of nearly critically sampled,
power-of-two time slices for a $1.4 - 1.4 \, M_{\odot}$ template extending from
$f_\mathrm{low} = 10 \, \mathrm{Hz}$ to $f_\mathrm{hi} = f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$
with a time-frequency structure given by ($\ref{eq:fgw})$. $f^s$ is the sample
rate of the time slice, $(t^{s+1}, t^s]$ are the boundaries in seconds
preceeding coalescence and \slicessamps\ is the number of samples in the
$s^{\mathrm{th}}$ filter.}
\end{table}
\end{comment}

@@

We note that the time slice decomposition in equation~(\ref{eq:time-slices}) is
manifestly orthogonal since the time slices are disjoint in time.  In the next
section we examine how to reduce the number of filters within each time slice
via singular value decomposition of the time sliced templates.

\subsection{Reducing the number of filters with the singular value
decomposition}

As described previously, the template banks are, by design, highly correlated.
It is possible to greatly reduce the number of filters required to achieve a
particular minimum match by designing an appropriate set of \SVD\ {\em
basis templates}.  A numerical technique based on the singular value
decomposition (\SVD) of inspiral template banks is demonstrated
in~\cite{Cannon:2010p10398}.  Similarly, the time sliced templates described
above can be approximated to arbitrary accuracy by expansion into a set of
\SVD\ \emph{basis templates}, $u_l^s[k]$, related to the original time
sliced templates through the \emph{reconstruction matrix},
$v_{il}^s\sigma_l^s$:
%
\begin{equation}
h_i^s[k] = \sum_{\mathclap{l=0}}^{\mathclap{M-1}} v_{il}^s \sigma_l^s u_l^s[k] \approx \sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s u_l^s[k].
\label{eq:svddecomp}
\end{equation}
%
\editorial{Say a word or two about \SVD\ tolerance?}%
%
The parameter $\numsvdtmps$ is the number of basis templates that are kept in
the approximation.  This determines the \SVD\ tolerance, which affects the
\SNR\ loss due to the approximation.  The authors of \cite{Cannon:2010p10398}
showed that high accuracy could be achieved with far fewer basis templates than
templates in the original template bank.  We find that when combined with the
time-slice decomposition, the number of basis templates \numsvdtmps\ is much
smaller than the original number of templates \numtmps\ and improves on
\cite{Cannon:2010p10398} by nearly an order of magnitude.  In the next section
we describe how we form our early warning detection statistic using the time-slice
decomposition and the \SVD.

@@

From the combination of transformations to the compact binary templates defined
in equation \eqref{eq:time-sliced-templates} and \eqref{eq:svddecomp} we define
the early-warning filter output accumulated up to time slice $s$ as,
%
% orthogonal decomposition filtering
%
\begin{equation}
	\rho_i^s [k] =%
		% Reconstruction
		\underbrace{
			\sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s
		}_\textrm{\clap{reconstruction}}
		% Orthogonal FIR filter
		\overbrace{
			\sum_{\mathclap{n=0}}^{\mathclap{N^s-1}} u_l^s[n] x^s[k-n]
		}^\textrm{\clap{orthogonal {\sc fir} filters}}
		% Plus ...
		+
		% Interpolation SNR
		\underbrace{
			\left(H^\uparrow \rho_i^{s+1}\right)[k]
		}_\textrm{\clap{{\sc snr} from previous time slices}}
\end{equation}
%
%
This quantity is related to the early warning \SNR{} by the normalization
of the partial filter defined by the present time slice, which may be computed
in advance. \editorial{Chad: are we happy with that statement?}
%
%
\begin{figure}[htbp]
	\includegraphics{figures/lloid-diagram.pdf}
	\caption{\label{fig:pipeline} Schematic of \lloid{} pipeline illustrating
signal flow.  Circles with arrows represent interpolation
\protect\includegraphics{figures/upsample-symbol.pdf} or decimation
\protect\includegraphics{figures/downsample-symbol.pdf}.  Circles with plus
signs represent summing junctions
\protect\includegraphics{figures/adder-symbol.pdf}.  Squares
\protect\includegraphics{figures/fir-symbol.pdf} stand for \fir{} filters.  Sample
rate decreases from the top of the diagram to the bottom.  In this diagram each
time slice contains three \fir\ filters that are linearly combined to produce
four output channels.  In a typical pipeline the number of \fir\ filters is
much less than the number of output channels.}
\end{figure}
%
%
The signal flow diagram in figure~\ref{fig:pipeline} illustrates how this
recursion relation can be realized as a multirate filter network with outputs for
each of the early warning \SNR{}s.

>>>>

In this section we describe a decomposition of the \CBC{} waveform space that
reduces time-domain (\TD) filtering cost sufficiently to allow for the
possibility of \earlywarning\ detection with modest computing requirements.  We
expand on the ideas of~\cite{Marion2004, Buskulic2010} that describe a
multiband decomposition of the compact binary parameter space that resulted in
a search with minutes latency in \LIGO{}'s S6 and Virgo's VSR3 joint science
runs~\cite{HugheyGWPAW2011}.  We combine this with the \SVD\ rank-reduction
method of~\cite{Cannon:2010p10398} that exploits the redundancy of
the template banks.

@@

Here we explore a method for reducing the computational cost of a \TD\ search
for compact binary coalescence.  We will give a truly zero-latency algorithm
that competes in terms of floating point operations per second with the
conventional \FD\ method, which requires a signficant latency in order to be
computationally cheap. Our method, \lloid\footnote{\lloid stands for \emph{Low
Latency Online Inspiral Detection} and was a code name for this procedure
throughout development. Although the name doesn't carry any specific
information about the algorithm, it became known within the gravitational-wave
community and has stuck.}%
involves two transformations of the template waveforms that produce a set of
orthogonal filters with far fewer coefficients than the original templates.

The first transformation is to chop the \TD{} templates up into \emph{time
slices}.  Since each template slice is disjoint in time, the resulting set for
a single template is orthogonal.  Given the chirp-like structure of the
templates, the early time slices have significantly lower bandwidth and can be
safely downsampled.  The downsampling removes a factor of $\sim$100 samples
from the early part of the waveform and allows the filters to be evaluated at
about $\sim$100 times lower rate.  This amounts to more than a factor of 10000
reduction in the floating-point operations per second required to filter the
early part of the waveform.  However, the resulting filters are still not
orthogonal across the mass parameter space, and are in fact highly redundant.
We use the \SVD{} to produce an orthogonal filter set
from the time-sliced templates~\cite{Cannon:2010p10398}.  We find that this
reduces the number of sample points of the filters by another factor of
$\sim$100.  The combined methods reduce the number of floating point operations
to the level where they are competitive with the conventional high-latency \FD\
matched filter approach.  In the remainder of this section we describe the
\lloid\ algorithm in detail and provide some basic computational cost scaling.

@@

The first step of our proposed method is to divide the templates into
time slices in a \TD{} analogue to the \FD{}
decomposition described in~\cite{Marion2004, Buskulic2010}.
A matched filter is constructed for each time
slice.  The outputs form an ensemble of partial \SNR{} streams.  By linearity,
these partial \SNR\ streams can be suitably time-delayed and summed to
reproduce the \SNR\ of the full template.  To wit,
%
\begin{equation}
\label{eq:time-slices}
h_{i}[k] = \sum_{s=0}^{S-1}
	\cases{%
		h_i^s[k] & for $t^s \leqslant k / f^0 < t^{s+1}$ \\
		0 & otherwise
	}
\end{equation}
\editorial{Drew: Should this be $h_{i}[k]$ or $\rho_{i}[k]$?}
%
for $S$ integers $\{f^0 t^s\}$ such that $0  = f^0 t^0 < f^0 t^1 < \cdots < f^0 t^S = N$.
We will show
in the next section that this, combined with the \SVD{},
is sufficient to enable a computationally efficient \TD{} search and
furthermore is an essential part of an \earlywarning\ detection scheme.

@@

The monotonic time-frequency relationship of equation~\eqref{eq:fgw} allows us
to choose time-slice boundaries that require substantially less bandwidth at
early times in the inspiral.  Our goal is to reduce the filtering cost of a
large fraction of the waveform by computing part of the convolution at a lower
sample rate.  Specifically we consider here time slice boundaries with the
smallest power-of-two sample rate that sub-critically samples the time-sliced
template.  The time slices consist of the $S$ intervals
$\left[t^0, t^1\right),\, \left[t^1, t^2\right),\, \dots,\, \left[t^{S-1}, t^S\right)$,
sampled at frequencies $f^0,\, f^1,\, \dots,\, f^\mathrm{S-1}$ where $f^0
\geqslant 2 f_\mathrm{hi}$ and $f^{S-1} \geqslant 2 f_\mathrm{low}$.  The time
sliced templates may be downsampled without aliasing, so we define them as
%
\begin{equation}
\label{eq:time-sliced-templates}
h_{i}^{s}[k] \equiv
	\cases{
		h_{i}\!\left[k\frac{f}{f^s}\right] & if $t^s \leqslant k/f^s < t^{s+1}$ \\
		0 & otherwise.
	}
\end{equation}
%
\begin{comment}
An example time-slice design satisfying these constraints for a $1.4 - 1.4 \,
M_{\odot}$ binary is shown in table~\ref{table:time_slices}.
%
\begin{table}[h!]
\begin{minipage}[c]{0.52\textwidth}
\centering
\vspace{0.8cm}
\includegraphics{time_slices.pdf}
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\centering
\input{time_slices.tex}
\end{minipage}
\caption{\label{table:time_slices} Example of nearly critically sampled,
power-of-two time slices for a $1.4 - 1.4 \, M_{\odot}$ template extending from
$f_\mathrm{low} = 10 \, \mathrm{Hz}$ to $f_\mathrm{hi} = f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$
with a time-frequency structure given by ($\ref{eq:fgw})$. $f^s$ is the sample
rate of the time slice, $(t^{s+1}, t^s]$ are the boundaries in seconds
preceeding coalescence and \slicessamps\ is the number of samples in the
$s^{\mathrm{th}}$ filter.}
\end{table}
\end{comment}

@@

We note that the time-slice decomposition in equation~(\ref{eq:time-slices}) is
manifestly orthogonal since the time slices are disjoint in time.  In the next
section we examine how to reduce the number of filters within each time slice
via \SVD{} of the time-sliced templates.

\subsection{Reducing the number of filters with the \SVD{}}

As described previously, the template banks are, by design, highly correlated.
It is possible to greatly reduce the number of filters required to achieve a
particular minimum match by designing an appropriate set of \SVD\ {\em
basis templates}, previously demonstrated
in~\cite{Cannon:2010p10398}.  Similarly, the time-sliced templates described
above can be approximated to arbitrary accuracy by expansion into a set of
\SVD\ \emph{basis templates}, $u_l^s[k]$, related to the original time
sliced templates through the \emph{reconstruction matrix},
$v_{il}^s\sigma_l^s$:
%
\begin{equation}
h_i^s[k] = \sum_{\mathclap{l=0}}^{\mathclap{M-1}} v_{il}^s \sigma_l^s u_l^s[k] \approx \sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s u_l^s[k].
\label{eq:svddecomp}
\end{equation}
%
\editorial{Say a word or two about \SVD\ tolerance?}%
%
The parameter $\numsvdtmps$ is the number of basis templates that are kept in
the approximation.  This determines the \SVD\ tolerance, which affects the
\SNR\ loss due to the approximation.  The authors of \cite{Cannon:2010p10398}
showed that high accuracy could be achieved with far fewer basis templates than
templates in the original template bank.  We find that when combined with the
time-slice decomposition, the number of basis templates \numsvdtmps\ is much
smaller than the original number of templates \numtmps\ and improves on
\cite{Cannon:2010p10398} by nearly an order of magnitude.  In the next section
we describe how we form our early-warning detection statistic using the time-slice
decomposition and the \SVD.

@@

From the combination of transformations to the compact binary templates defined
in equation \eqref{eq:time-sliced-templates} and \eqref{eq:svddecomp} we define
the early-warning filter output accumulated up to time slice $s$ as,
%
% orthogonal decomposition filtering
%
\begin{equation}
	\rho_i^s [k] =%
		% Reconstruction
		\underbrace{
			\sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s
		}_\textrm{\clap{reconstruction}}
		% Orthogonal FIR filter
		\overbrace{
			\sum_{\mathclap{n=0}}^{\mathclap{N^s-1}} u_l^s[n] x^s[k-n]
		}^\textrm{\clap{orthogonal {\sc fir} filters}}
		% Plus ...
		+
		% Interpolation SNR
		\underbrace{
			\left(H^\uparrow \rho_i^{s+1}\right)[k]
		}_\textrm{\clap{{\sc snr} from previous time slices}}
\end{equation}
%
%
This quantity is related to the early-warning \SNR{} by the normalization
of the partial filter defined by the present time slice, which may be computed
in advance. \editorial{Chad: are we happy with that statement?}
%
%
\begin{figure}[htbp]
	\includegraphics{figures/lloid-diagram.pdf}
	\caption{\label{fig:pipeline} Schematic of \lloid{} pipeline illustrating
signal flow.  Circles with arrows represent interpolation
\protect\includegraphics{figures/upsample-symbol.pdf} or decimation
\protect\includegraphics{figures/downsample-symbol.pdf}.  Circles with plus
signs represent summing junctions
\protect\includegraphics{figures/adder-symbol.pdf}.  Squares
\protect\includegraphics{figures/fir-symbol.pdf} stand for \fir{} filters.  Sample
rate decreases from the top of the diagram to the bottom.  In this diagram each
time slice contains three \fir\ filters that are linearly combined to produce
four output channels.  In a typical pipeline the number of \fir\ filters is
much less than the number of output channels.}
\end{figure}
%
%
The signal flow diagram in figure~\ref{fig:pipeline} illustrates how this
recursion relation can be realized as a multirate filter network with outputs for
each of the early-warning \SNR{}s.
